\section{Discussion}

For a molecular ecologist, the value of VariantFlow is simple to state: the
routine questions you ask of a large cohort VCF---how much missing data per
site? which sites pass a quality cut? what are diversity and differentiation
across windows?---come back while you are still at the keyboard rather than
after a coffee break or an overnight batch job, and they come back with exactly
the answer the established tools would have given. The reason is the book image
that runs through this paper: VariantFlow reads your question first, opens only
the labelled sections of each record that could hold the answer, and leaves the
rest closed---so it touches a fraction of the ink and returns the very same
number a tool that read every field would have produced, with the skimmed-past
records handed back untouched. That is why the wait
shrinks without the science changing: this field-selective, single-pass
approach (``selective execution'') does less redundant reading, never a corner
cut on the answer.

VariantFlow is a companion to the mature tools, not a replacement for them, and
it helps most on a specific shelf of tasks: reading through native VCF and its
compressed form (BGZF, the \texttt{.vcf.gz} files you already work with) to
apply selective filters; filters that use the file's index---the book's table
of contents---to jump straight to the relevant region and skip most of the
file; the population summaries VCFtools already supports; and the
export-once, query-many pattern that writes a compact table once (in a compact
table format called Parquet) and then answers many questions from it. It does
not claim universal replacement of bcftools, HTSlib, GATK, or VCFtools; a
summary table shipped with the software lists where it beats, matches, or
complements each existing tool.

These tools occupy complementary niches, and you can choose by the job in front
of you: \texttt{bcftools} remains the reference for general filtering,
normalization, and format conversion; VCFtools for classic population-genetics
file operations; DuckDB (a lightweight database) for free-form questions in SQL
(a query language) over tables you have already written out; and scikit-allel
for matrix- and haplotype-based statistics (selection scans, principal-component
analysis, $D$/$f$-statistics) that need every sample's genotypes loaded into the
computer's working memory simultaneously---every section of every book open
together, which is exactly what VariantFlow deliberately avoids. VariantFlow fills the gap between them: fast,
field-selective filtering and streamable site- and window-wise statistics on
large cohorts, with export-once handoff to DuckDB and explicit deferral to
scikit-allel for matrix-wide analyses (Supplementary Table~S5). The practical
rule is to reach for VariantFlow when interactive turnaround on large VCFs on
ordinary desktop and laptop CPUs is the priority, and for the established tools
when their full generality is required.

GPU-accelerated pipelines such as NVIDIA Clara Parabricks target variant
calling and the upstream processing that produces a VCF in the first place,
using massive parallelism \citep{parabricks}. VariantFlow addresses a different
bottleneck: the post-calling step of selecting fields on the ordinary desktop
and laptop CPUs most labs already have. The two approaches are complementary---one
speeds up making the books, the other speeds up reading only the passages you
need from them.

If reading only the sections a question names becomes common in post-calling
tools, routine workflows on large cohorts may shift from batch to interactive
timescales, the export-once pattern can reduce redundant reading and rewriting
of data, and tool authors may re-examine which fields each pipeline stage
actually needs to open. We anticipate the most immediate value for groups that
repeatedly process large cohort VCFs on ordinary hardware: population and
conservation/ecological geneticists computing diversity, differentiation, and
linkage disequilibrium (LD); plant and animal breeding programmes screening
large panels; large-cohort
human-genomics projects performing quality control and missingness or
allele-frequency summaries; and pipeline developers who need a fast, scriptable
filtering and export layer. For these communities VariantFlow's distinctive
value is bringing whole-cohort post-calling summaries from batch to interactive
timescales without a cluster, while preserving exact correctness against
established tools.

Because the whole promise rests on getting the same answer, correctness is
checked directly rather than assumed. VariantFlow's output is validated by
byte-identical matching against VCFtools---the result agrees to the last
character---and cross-checked with scikit-allel \citep{scikitallel}. Missing
data is the case where a diversity estimate can quietly go wrong, so it is
tested explicitly: for diversity under missing data, VariantFlow implements the
unbiased per-site $\pi$ and $d_{XY}$ estimators of pixy \citep{pixy2021}; on an
all-sites cohort with 15\% missing genotypes its pairwise counts are identical
to pixy across every window (Supplementary Table~S4), delivered as a compiled
selective-execution pass that opens only the sections those estimators depend
on. We are equally plain about where VariantFlow does not lead: it reads only a
limited set of per-sample FORMAT fields, so questions that depend on FORMAT keys
it does not yet open must still go to the established tools; its
Hardy--Weinberg equilibrium (HWE)
calculation slightly trails VCFtools, and its BGZF filtering does not outperform
bcftools on many-sample files. Future work will broaden expression
compatibility.
