\section{Algorithm and implementation}

Start with the wait you already know. You have a chr22 cohort with thousands of
samples, and you only want to keep the sites where the quality score
(\texttt{QUAL}) is above 30. A conventional tool still reads out every genotype
and read-depth value for every sample at every site---faithfully
re-transcribing millions of numbers it will immediately throw away---before it
can answer your one narrow question. On a multi-thousand-sample cohort, that
transcription of data you never asked for, not the biology, is what you wait
for.

VariantFlow turns only to the sections your question names and reads only the
words it needs on each. In implementation terms, this is the core design
principle of selective execution: a query is compiled to the minimal set of
fields it touches, and only those fields are decoded---that is, opened
(Figure~1)---in contrast to the conventional approach of decoding every field
of every record. It reads each record where it already sits, glancing at the
raw text without first copying the whole record into memory---what the engine
calls treating each VCF record as a borrowed byte-slice view (reading the
label without copying the book out). Site fields, selected INFO values, FORMAT
keys, and sample values are found by scanning straight to just the labelled
sections the question names, rather than first turning every column of every
record into text it may never use (byte scanning rather than materializing all
columns as strings). To know which sections to open, VariantFlow first turns
your query into a plain checklist of exactly which fields the answer could
depend on---a typed predicate tree, where a \emph{predicate} is simply the
condition to check, such as \texttt{QUAL>30}---and then asks each record only
for the fields on that checklist. The work it does therefore stays close to
what you actually asked for. And because the sections it skips are, by
construction, ones the answer never depended on, the number you get is
identical to a tool that read every field---verified here as byte-identical
against VCFtools~0.1.17, and cross-checked against scikit-allel and pixy. The
speed comes from doing less redundant reading, never from cutting a corner on
the science.

For a filter that touches many samples, VariantFlow reads the label
\texttt{GT:DP:GQ:AD} once, notes where the single key it needs (say
\texttt{DP}, read depth) sits, and then reads just that one value by its
position across every sample---instead of re-reading the whole label for each
of the thousands of samples. In implementation terms, it scans the FORMAT
schema once per record, caches the requested key index, and reads sample values
by index. And in the mode that keeps each record's original text, the records
it skims past pass through unchanged: native filtering writes passing records
in input order and preserves their original text in the line-preserving path.

The same read-only-what-you-need idea reaches down to the plumbing. Compressed
files are read through a native reader for the standard compressed VCF format
(BGZF), using parallel workers capped automatically to the machine (auto-capped
worker threads). When a condition is expensive to check, records are evaluated
in ordered batches so the output still comes back in the original order. For a
filter you run again and again that rejects most records, VariantFlow can read
a short table of contents written alongside the file---a \texttt{.vfi} sidecar
index---that summarises each block of records (its genomic range, QUAL, FILTER,
and selected INFO summaries), so it can leave an entire block closed whenever
that summary proves no record inside it could possibly pass the filter. Where
available, optional paths through the standard genomics C library HTSlib add
reading of BCF (the binary form of VCF), reading only an indexed region of the
file, and BGZF output.

VariantFlow also serves population-genetics and export workflows. Where the
data are two-copy, two-allele (diploid biallelic) genotypes, the VCFtools-style
commands work from compact genotype summaries, preserving VCFtools-compatible
output for frequency, missingness, HWE, LD, Fst, and related population
summaries. Parquet~\citep{parquet} export writes the data once as typed
columns---a layout that lets tools such as DuckDB \citep{duckdb} re-run many
analyses over the same data quickly.

The command-line interface follows the small-tool style common in genomics,
organised as a small set of composable subcommands (Figure~2):
{\small
\texttt{variantflow filter input.vcf.gz --where "QUAL>30 \&\& ANY(FORMAT/DP>20)"}\\
\texttt{variantflow fst cohort.vcf.gz --pop pop1.txt --pop pop2.txt -o fst.tsv}
}
