\section{Introduction}

Genomic variant data is accumulating rapidly. Population-scale projects now
routinely produce VCF files with thousands of samples and millions of variant
sites per chromosome \citep{vcftools}. Post-calling workflows---quality
filtering, cohort subsetting, population summaries, exporting a table for
downstream analysis---scan these files repeatedly, and each pass typically
parses the full record even when the operation depends on only QUAL (the
site's quality score), one INFO tag (one site-level annotation), or a single
FORMAT key (one per-sample value such as a genotype or a read depth). On a
multi-thousand-sample cohort, parsing unused FORMAT vectors---the per-sample
genotype-and-depth entries---and sample values can dominate runtime without
changing the result.

Mature tools---bcftools, HTSlib, VCFtools---form the reliable
infrastructure on which larger pipelines rely
\citep{bcftools,htslib}. The VCF is an exchange format, and it is worth
being concrete about what that means for the biologist at the keyboard. A
single VCF record is one long text row that packs together every property of
one variant: its genomic position, its quality score, its filter status, and
then, for every sample in the cohort, that sample's genotype, read depths,
and a long list of annotations. Everything about the site is written out
side by side in the same row. That is a deliberate strength---the format was
built so any program doing any analysis can find whatever it might need, which
is exactly why VCF is the shared currency of variant data. But it carries a
hidden tax. Because every property sits in the same row, a program that opens
the file must read past all of it even to answer a question about one column;
on a cohort with thousands of samples, that means megabytes of unwanted
genotype and depth text are decoded for every question asked. When your actual
question is narrow---``keep sites where QUAL $>$ 30,'' or ``how much missing
data per site?''---the conventional tool still transcribes every record in
full, sample by sample, before it can answer, and on a
multi-thousand-sample cohort that transcription of unread sections---not the
biology---is what dominates the wait. This is nobody's mistake: an exchange
format built to hold everything for everyone is not always the fastest way to
lay out the data for the one task in front of you.

Consider how you read a reference book when you already know your question:
instead of copying out every page in full, you read your question first, work
out the few passages that could hold the answer, turn only to those, and read
only the words you need on each. Every variant record is like a still-closed
book with many labelled sections---QUAL, each INFO tag, and, for every one of
thousands of samples, a FORMAT entry such as GT:DP:GQ:AD (the genotype, its
read depth, its genotype quality, and the per-allele depths); a conventional
tool copies out every section of every book before it can answer, whereas
VariantFlow reads only the sections your question names and leaves the rest
closed. This is what \emph{selective execution} means. In honest terms that
mirror the implementation: VariantFlow first reads your query---a filter, a
population statistic, or an export---and compiles it into a checklist built
from your question of exactly which fields the answer could possibly depend
on---organised by field type so the tool knows how to read each one (in the
implementation, a typed predicate tree); then, going record by record and
reading through each record only once (a single streaming pass), it scans
straight to just the fields on the checklist---reading the labels where they
sit in the file rather than first copying the whole record out into memory (it
reads the raw bytes in place, a borrowed view rather than a copy); it reads
only those fields and skips everything else, and, where the task allows, hands
the original record text back unchanged.
This removes the exact pain point named above: the wait came from
transcribing unread sections, so reading only the sections and words your
question needs removes the wait without touching the answer. The fields it
skips are, by construction, ones the answer never depended on, so the number
you get is exactly the same as if it had copied out every page---it just
touched a fraction of the ink. Concretely, VariantFlow parses only the fields
required by the current filter, statistic, or exported selection of columns
(predicate, statistic, or projection); preserves original records where
possible; and tracks every performance claim against
correctness-matched benchmarks. It complements bcftools, HTSlib, VCFtools,
and GATK~\citep{gatk} rather than replacing them.
