\section{Benchmark evidence}

Recall the book image that motivates the tool: VariantFlow opens only the
labelled sections of a record its question names, and leaves the rest closed.
The measurements below ask a plain empirical question about that strategy: when
VariantFlow opens fewer sections, how much time does a working biologist get
back, and is the answer still exactly the one a tool that read every field
would have produced?

All measurements are tied to tracked benchmark reports with exact commands,
competitor versions, correctness checks, and caveats (Table~1).
VariantFlow reports a speedup only when the benchmark row records a
correctness comparison; rows that trail a competitor remain optimization
targets rather than claims. In book terms, a number is only claimed once we have
confirmed that reading fewer sections returned the same answer as reading them
all.

On the DDBJ CHM13 human chr22 cohort (3\,715 samples), when the question names
per-sample FORMAT sections, VariantFlow turns only to those sections and still
reaches the same records bcftools does, at 4.74--17.78x faster on bounded
record tiers (a FORMAT-rich \emph{predicate} is the checklist of fields
VariantFlow builds from your question). A companion index (a \texttt{.vfi} file) records, for each block
of records, whether any record in it could possibly match---so VariantFlow can
skip whole runs of books unopened. For \texttt{FILTER == "PASS"} queries where
most blocks can be ruled out in advance (high-skip predicates, where the block
metadata proves no record can pass), this reached 123--273x; when few blocks can
be ruled out, VariantFlow simply reads through every record one by one
(low-skip predicates fall back to a sequential scan). Here the biggest
gains come exactly where the book image predicts them: when the index can prove
in advance that a whole run of records cannot answer your question, those books
are never opened at all, and a filtering pass that touched every record collapses
to one that skips past most of them.

Population-genetics summaries on the full IGSR 3\,202-sample biallelic cohort
show that selective execution scales to whole-chromosome data. Missingness
measured 4.78x faster than VCFtools on chr22 (1.07M records) and 3.67x on
chr1 (6.47M records), with a constant 9~MB peak memory footprint (peak RSS)
regardless of chromosome
size. In everyday terms, a per-site missing-data summary returns several times
faster than VCFtools, and the constant 9~MB peak RSS
means a whole-chromosome scan runs on a laptop rather than demanding a
large-memory node---the wait shrinks because VariantFlow reads only the one
section each site's answer depends on and streams past the genotype vectors it
would otherwise transcribe and discard. All population-genetics outputs were
verified by byte-identical diff against
VCFtools~0.1.17 on the same inputs (Supplementary Table~S1); an independent
cross-check with scikit-allel confirmed numerical agreement---the same answer,
verified byte-identical, not merely a close one.
Linkage disequilibrium (LD)---which sites are inherited together---measured
9.88x on a 925{,}000-record dataset (925k-record tier). VariantFlow can also
rewrite the file once into a column-organised layout (Parquet) so later questions
read straight to their column---and querying that with a fast query engine
(DuckDB) measured 1.19--19.20x, averaged over many repeated queries (amortized):
you pay once to turn the pages into a columnar layout, then every later question
reads straight to its column instead of reopening the whole book.

VariantFlow implements the population-genetics statistics that are computable
in a single streaming pass over sites or windows---that is, the questions that
can be answered by reading through the book once, turning only to the sections
each one names---allele frequency,
missingness, heterozygosity, Hardy--Weinberg, $\pi$, missing-data-aware
$\pi$/$d_{XY}$, Tajima's $D$, the site-frequency spectrum, $F_{ST}$, and
LD (Figure~1). The site-frequency spectrum (\texttt{sfs})
reproduces scikit-allel's folded and unfolded spectra exactly. Statistics that
require the full genotype or haplotype matrix in memory---selection scans,
principal-component analysis, and $D$/$f$-statistics---need every section of
every book open at once, are outside this
streaming scope, and are deferred to scikit-allel; Supplementary Table~S5 maps
the full correspondence. VariantFlow is a companion that opens fewer sections
for the questions it covers, not a replacement for the mature tools that hold
everything for everyone.
