\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{xcolor}
\usepackage{amssymb}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{lineno}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc}

\definecolor{vfblue}{HTML}{1F4E79}
\definecolor{vfteal}{HTML}{2A7F73}
\definecolor{vflight}{HTML}{EEF5F8}

\hypersetup{
  colorlinks=true,
  linkcolor=vfblue,
  citecolor=vfblue,
  urlcolor=vfblue
}
\setlength{\emergencystretch}{2em}

\title{VariantFlow: a selective-execution engine for efficient population genomic computation on large variant datasets}
\author{
Ehsan Estaji\textsuperscript{1} \qquad
Jian-Feng Mao\textsuperscript{1,$\dag$} \\[6pt]
\small \textsuperscript{1}Ume{\aa} Plant Science Centre,
Department of Plant Physiology, Ume{\aa} University,
SE-901\,87 Ume{\aa}, Sweden \\[4pt]
\small \textsuperscript{$\dag$}Correspondence:
       \href{mailto:jianfeng.mao@umu.se}{\texttt{jianfeng.mao@umu.se}}
}
\date{}

\begin{document}
\linenumbers
\maketitle
\newpage

\begin{abstract}
\noindent
Population-scale sequencing projects now produce variant call sets with
thousands of samples and millions of sites, and post-calling analysis of these
files has become a recurring computational bottleneck. Because the Variant Call
Format stores every field of every record together, a tool answering a
field-limited question still parses the unused annotations, FORMAT blocks, and
per-sample values, which costs time without changing the result. We present
VariantFlow, a command-line engine whose central design principle is selective
execution, decoding only the fields required by the current filter, statistic,
or projection and preserving original records where possible. On
correctness-matched benchmarks using the 1000 Genomes 3\,202-sample
high-coverage dataset, VariantFlow computed missingness 3.67--4.78x faster than
VCFtools from single-chromosome to whole-genome scale at a constant 8.5~MB of
memory, filtered FORMAT-rich predicates 4.74--17.78x faster than bcftools,
evaluated index-assisted high-skip predicates 123--273x faster, computed
linkage disequilibrium 9.88x faster, and served export-once DuckDB queries
1.19--19.20x faster; every reported statistic was byte-identical to VCFtools
and, where applicable, to scikit-allel and pixy. By decoding only the fields
each operation requires, VariantFlow brings large-cohort post-calling analysis
within reach of interactive, exploratory work on commodity hardware. It
complements rather than replaces bcftools, HTSlib, GATK, and VCFtools, and we
expect it to be of immediate value in population and conservation genomics,
plant and animal breeding, large-cohort human-genomics quality control, and
bioinformatics pipeline development. VariantFlow is open source (Rust,
MIT OR Apache-2.0) at \url{https://github.com/ehsanestaji/VariantFlow}, and
version~1.5.0 is archived at
\href{https://doi.org/10.5281/zenodo.21198171}{doi:10.5281/zenodo.21198171}.

\vspace{0.6em}
\noindent\textbf{Keywords:} population genomics, variant call format,
selective execution, missing data, nucleotide diversity, bioinformatics
software, Rust
\end{abstract}
\newpage

\input{sections/introduction}
\input{sections/implementation}
\input{sections/results}
\input{sections/discussion}

\section*{Author contributions}
E.E.\ conceived the tool, designed and implemented the software, conducted
benchmarks, and wrote the manuscript. J.-F.M.\ supervised the project,
coordinated external feedback, and reviewed the manuscript. Both authors read
and approved the final manuscript.

\section*{Funding}
This work was partially supported by the Wallenberg Initiatives in Forest
Research (WIFORCE) funded by the Knut and Alice Wallenberg Foundation.

\section*{Conflict of interest}
The authors declare no competing interests.

\section*{AI usage disclosure}
AI coding assistants were used for code-review support and test
scaffolding. The authors remain responsible for all scientific claims.

\section*{Acknowledgements}
Computing resources provided by SNIC/HPC2N at Ume{\aa} University.

\bibliographystyle{apalike}
\bibliography{references}

% ======================================================================
\section*{Data Accessibility and Benefit-Sharing Statement}

\noindent\textbf{Data Accessibility Statement.} VariantFlow's source code,
build instructions, and all reproducibility materials (exact commands,
benchmark reports, and correctness-validation scripts) are openly available
at \url{https://github.com/ehsanestaji/VariantFlow} under the MIT OR
Apache-2.0 license, and the version described in this manuscript (v1.5.0) is
permanently archived at Zenodo, \href{https://doi.org/10.5281/zenodo.21198171}{doi:10.5281/zenodo.21198171}.
No new biological samples or sequence data were generated for this study.
All benchmarks reuse publicly available, previously deposited variant call
sets from the 1000 Genomes Project high-coverage release
(International Genome Sample Resource), used under their existing public
data-access terms.

\vspace{0.6em}
\noindent\textbf{Benefit-Sharing Statement.} This study did not involve the
collection of new biological specimens, field sampling, or engagement with
Indigenous or local communities; it reuses previously published, de-identified
public human genomic datasets. Benefit-sharing provisions are therefore not
applicable to this work.

% ======================================================================
% Figure/table legends and supplementary list, after the references.
\clearpage
\section*{Figure and table legends}

\noindent\textbf{Table~1. Correctness-matched benchmarks from tracked
reports.} Each row records exact commands, competitor version, correctness
checks, and peak resident memory. IGSR rows use the 1000~Genomes 3202-sample
high-coverage dataset. \checkmark~$=$ byte-identical output vs.\ baseline.
Baselines: \texttt{bcftools filter} (filtering), VCFtools~0.1.17 (population
summaries), repeated scans (DuckDB). See Supplementary Table~S1 for cross-tool
validation including scikit-allel.

\vspace{0.9em}
\noindent\textbf{Figure~1. Selective execution versus conventional
full-record parsing.} (a) Conventional tools decode every field of every VCF
record before applying an operation, so most decoded bytes are discarded.
(b) VariantFlow compiles each query to the minimal set of fields it touches and
decodes only those, preserving the original record. The lower band lists the
population-genetics statistics VariantFlow computes in a single streaming pass;
matrix-wide and haplotype-based statistics are deferred to scikit-allel
(Supplementary Table~S5).

\vspace{0.9em}
\noindent\textbf{Figure~2. The VariantFlow command tree.} Subcommands are
grouped into filtering/indexing/I-O and population-genetics families; each leaf
gives the command and a one-line description of its function.

\section*{Supplementary material}
The supplementary document (\texttt{supplementary.pdf}) contains the following
items; each is self-contained and begins on its own page:

\begin{itemize}\setlength{\itemsep}{2pt}
\item Supplementary Table S1 --- cross-tool correctness validation against VCFtools and scikit-allel.
\item Supplementary Table S2 --- benchmark system configuration.
\item Supplementary Table S3 --- full benchmark results with timing and memory.
\item Supplementary Table S4 --- missing-data validation against pixy.
\item Supplementary Table S5 --- statistics coverage relative to scikit-allel, grouped by computational class.
\item Supplementary Note S1 --- missing-data-aware $\pi$ and $d_{XY}$.
\item Supplementary Note S2 --- correctness validation methodology.
\item Supplementary User Guide --- an end-to-end population-genetics analysis of 1000~Genomes chromosome~22, from installation to interpretation (also available online).
\end{itemize}

% ======================================================================
% Display items: title-only, one per page (page break after supplementary list).
\clearpage
\begin{center}\textbf{Table 1}\end{center}
\vspace{0.5em}
\begin{center}
\small
\begin{tabularx}{\linewidth}{p{0.18\linewidth}p{0.18\linewidth}Xp{0.20\linewidth}c}
\toprule
Workflow & Dataset & Speedup & RSS, VariantFlow\,/\,baseline & Match \\
\midrule
Human FORMAT filters & CHM13 chr22, 3\,715 sam. & 4.74--17.78x & 11\,/\,9 MB & \checkmark \\
Indexed \texttt{PASS} & IGSR chr22, skip chunks & 123--273x & 11\,/\,10 MB & \checkmark \\
Indexed AF pred. & IGSR chr22, 256-rec & 4.70--4.93x & 11\,/\,10 MB & \checkmark \\
Missingness & IGSR chr22 full, 1.07M & 4.78x & 8.5\,/\,4.1 MB & \checkmark \\
Missingness & IGSR chr1 full, 6.47M & 3.67x & 8.5\,/\,4.1 MB & \checkmark \\
LD ($r^{2}$) & IGSR chr22, 925k & 9.88x & 11.4\,/\,11.4 MB & \checkmark \\
DuckDB amortized & Parquet export-once & 1.19--19.20x & --- & \checkmark \\
\bottomrule
\end{tabularx}
\end{center}

\clearpage
\begin{center}\textbf{Figure 1}\end{center}
\vspace{0.5em}
\begin{center}
\input{figures/figure1_architecture}
\end{center}

\clearpage
\begin{center}\textbf{Figure 2}\end{center}
\vspace{0.5em}
\begin{center}
\input{figures/figure2_commandtree}
\end{center}

\end{document}
