\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xcolor}

\definecolor{vfblue}{HTML}{1F4E79}
\hypersetup{colorlinks=true,linkcolor=vfblue,citecolor=vfblue,urlcolor=vfblue}

\newcommand{\yes}{\checkmark}
\newcommand{\no}{--}

\title{Supplementary Materials\\[2pt]
\large VariantFlow: a selective-execution engine for efficient population\\
genomic computation on large variant datasets}
\author{Ehsan Estaji and Jian-Feng Mao}
\date{}

\begin{document}
\maketitle

\noindent This document is self-contained: each supplementary item (table or
note) begins on its own page and restates the context needed to read it
independently of the main text.

\vspace{1em}
\noindent\textbf{Contents}
\begin{itemize}\itemsep2pt
  \item Table S1 --- Cross-tool correctness validation (VariantFlow vs.\ VCFtools and scikit-allel).
  \item Table S2 --- Benchmark system configuration.
  \item Table S3 --- Full benchmark results with timing and memory.
  \item Table S4 --- Missing-data validation against pixy.
  \item Table S5 --- Statistics coverage of VariantFlow relative to scikit-allel, grouped by computational class.
  \item Note S1 --- Missing-data-aware $\pi$ and $d_{XY}$.
  \item Note S2 --- Correctness validation methodology.
\end{itemize}

% ======================================================================
\newpage
\section*{Table S1: Cross-tool correctness validation}

VariantFlow outputs were compared against VCFtools~0.1.17 and scikit-allel~1.3.13
on the IGSR 1000 Genomes chr22 high-coverage dataset (1,073,369 biallelic
SNV records, 3,202 diploid samples). Correctness is assessed by exact
output match (integer fields) or maximum absolute difference (floating-point
fields). VariantFlow is judged correct only where the difference is zero (or
within machine epsilon for floating-point).

\begin{table}[ht]
\centering
\small
\begin{tabularx}{\linewidth}{lXccc}
\toprule
Statistic & Comparison tool & Records & Max $|\Delta|$ & Exact? \\
\midrule
Per-individual missingness (F\_MISS) & VCFtools & 3,202 & 0.0 & Yes \\
Per-site missingness (F\_MISS) & VCFtools & 1,073,369 & 0.0 & Yes \\
Allele frequency & VCFtools & 1,073,369 & 0.0 & Yes \\
Nucleotide diversity ($\pi$) & VCFtools & 1,073,369 & 0.0 & Yes \\
LD ($r^{2}$) & VCFtools & 925,000 & 0.0 & Yes \\
Weir--Cockerham $F_{ST}$ & VCFtools & 1,073,369 & 0.0 & Yes \\
Per-individual missingness & scikit-allel & 3,202 & $<10^{-15}$ & Yes$^{*}$ \\
Allele frequency & scikit-allel & 1,073,369 & $<10^{-15}$ & Yes$^{*}$ \\
\bottomrule
\end{tabularx}
\vspace{4pt}

\footnotesize{$^{*}$Floating-point values agree within machine epsilon
($\approx 2.2 \times 10^{-16}$). Integer counts are identical.}
\end{table}

% ======================================================================
\newpage
\section*{Table S2: Benchmark system configuration}

All benchmarks reported in the main text and in Table~S3 were run on the
single configuration below. Timing is wall-clock, single-threaded; peak
memory is peak resident set size (RSS).

\begin{table}[ht]
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\toprule
Parameter & Value \\
\midrule
Hardware & Apple M5 Max, 14-core CPU, 128~GB unified memory \\
OS & macOS 15.x (arm64) \\
VariantFlow & v1.5.0 (Zenodo doi:10.5281/zenodo.21198171) \\
bcftools & 1.21 (htslib 1.21) \\
VCFtools & 0.1.17 \\
scikit-allel & 1.3.13 \\
pixy & 2.2.1 \\
Input data & IGSR 1000 Genomes high-coverage (GRCh38) \\
chr22 & 1,073,369 biallelic records, 3,202 samples \\
chr1 & 6,468,094 biallelic records, 3,202 samples \\
pixy validation set & 5,000 all-sites records, 40 samples, 2 populations, 15\% missing \\
Memory metric & Peak RSS via \texttt{resource.getrusage(RUSAGE\_CHILDREN)} \\
Timing & Wall-clock seconds, single-threaded execution \\
\bottomrule
\end{tabularx}
\end{table}

% ======================================================================
\newpage
\section*{Table S3: Full benchmark results with timing and memory}

Per-workflow wall-clock time and peak memory for VariantFlow and the baseline
tool, on the system described in Table~S2. Baseline is VCFtools~0.1.17 for all
population-genetics rows. Every row was correctness-checked before timing
(Table~S1).

\begin{table}[ht]
\centering
\small
\begin{tabularx}{\linewidth}{llrrrr}
\toprule
Workflow & Dataset & VariantFlow (s) & Baseline (s) & VariantFlow RSS (MB) & Baseline RSS (MB) \\
\midrule
Missingness & chr22 full & 29.6 & 141.6 & 8.5 & 4.1 \\
Missingness & chr1 full & 207.7 & 762.0 & 8.5 & 4.1 \\
LD ($r^{2}$) & chr22 (925k) & 77.0 & 791.3 & 11.4 & 11.4 \\
Frequency & chr22 (925k) & 8.2 & 30.4 & 11.4 & 4.2 \\
$\pi$ (per-site) & chr22 (925k) & 9.1 & 33.7 & 11.4 & 4.3 \\
$F_{ST}$ (Weir--Cockerham) & chr22 (925k) & 12.3 & 16.9 & 11.4 & 4.8 \\
\bottomrule
\end{tabularx}
\vspace{4pt}

\footnotesize{All runs single-threaded on macOS arm64. RSS = peak resident
set size. Speedups quoted in the main text are baseline/VariantFlow wall-clock ratios.}
\end{table}

% ======================================================================
\newpage
\section*{Table S4: Missing-data validation against pixy}

The pixy tool \citep{pixy2021} provides unbiased estimators for nucleotide
diversity ($\pi$) and between-population divergence ($d_{XY}$) that account
for missing genotypes by counting only non-missing pairwise comparisons at
each site, aggregated as a ratio of sums across all sites in a window
(including invariant sites). VariantFlow implements the identical estimator
in its \texttt{pixy} subcommand (Note~S1). To validate this, we generated a
deterministic all-sites VCF (5,000 sites, 40 diploid samples, two
populations) with 15\% of genotypes set to missing, and computed windowed
$\pi$ and $d_{XY}$ (1\,kb windows) with both tools.

\begin{table}[ht]
\centering
\small
\begin{tabularx}{\linewidth}{lXccc}
\toprule
Statistic & Comparison & Windows & Count match & Max $|\Delta|$ \\
\midrule
$\pi$ (per population)   & pixy 2.2.1 & 10 & Exact & $<5\times10^{-9}$ \\
$d_{XY}$ (per pop.\ pair) & pixy 2.2.1 &  5 & Exact & $<5\times10^{-9}$ \\
\bottomrule
\end{tabularx}
\vspace{4pt}

\footnotesize{Integer pairwise counts (\texttt{count\_diffs},
\texttt{count\_comparisons}) are byte-identical to pixy across every window;
the residual $|\Delta|$ in the averaged ratio reflects only the eight-decimal
print precision of VariantFlow. The comparison is performed on data
containing missing genotypes, the regime pixy was designed for.}
\end{table}

% ======================================================================
\newpage
\section*{Table S5: Statistics coverage relative to scikit-allel}

Population-genetics statistics grouped by computational class. The organizing
rule is the memory-access pattern each statistic requires. \textbf{Group A}
statistics are computable in a single streaming pass over sites or windows and
are implemented by VariantFlow's selective-execution engine; overlapping
statistics are benchmarked against scikit-allel~\citep{scikitallel} and/or
VCFtools. \textbf{Group B} statistics require the full genotype or phased
haplotype matrix to be held in memory simultaneously; these are the domain of
matrix-oriented libraries such as scikit-allel, and VariantFlow defers to them
by design rather than reimplementing them. This delineation reflects
VariantFlow's scope: it complements, and does not replace, scikit-allel.

\begin{table}[ht]
\centering
\small
\begin{tabularx}{\linewidth}{Xllc}
\toprule
Statistic & VariantFlow & scikit-allel name & Validated \\
\midrule
\multicolumn{4}{l}{\textit{Group A --- site-wise / window-wise (streaming; VariantFlow's domain)}}\\
\midrule
Allele frequency / counts & \texttt{freq} & \texttt{allele\_frequencies} & \yes\ (VCFtools, s-a) \\
Genotype missingness & \texttt{missingness} & (genotype masks) & \yes\ (VCFtools, s-a) \\
Observed/expected heterozygosity & \texttt{het} & \texttt{heterozygosity\_*} & VCFtools \\
Hardy--Weinberg equilibrium & \texttt{hardy} & \no & VCFtools \\
Nucleotide diversity $\pi$ & \texttt{pi} & \texttt{sequence\_diversity} & \yes\ (VCFtools) \\
Missing-data-aware $\pi,\,d_{XY}$ & \texttt{pixy} & (pixy) & \yes\ (pixy, exact) \\
Tajima's $D$ & \texttt{tajima-d} & \texttt{tajima\_d} & VCFtools \\
Watterson's $\theta$ & \texttt{tajima-d}$^{\dag}$ & \texttt{watterson\_theta} & partial \\
Site frequency spectrum (folded/unfolded) & \texttt{sfs} & \texttt{sfs}, \texttt{sfs\_folded} & \yes\ (s-a, exact) \\
$F_{ST}$ (Weir--Cockerham, Hudson) & \texttt{fst} & \texttt{*\_fst} & \yes\ (VCFtools) \\
Linkage disequilibrium $r^{2}$ & \texttt{ld} & \texttt{rogers\_huff\_r} & \yes\ (VCFtools) \\
\midrule
\multicolumn{4}{l}{\textit{Group B --- matrix-wide / haplotype-based (defer to scikit-allel by design)}}\\
\midrule
Joint / scaled SFS (2D) & \no & \texttt{joint\_sfs}, \texttt{sfs\_scaled} & --- \\
Selection scans (EHH, iHS, XP-EHH, nSL) & \no & \texttt{ehh}, \texttt{ihs}, \dots & --- \\
Garud's $H$, haplotype diversity & \no & \texttt{garud\_h}, \dots & --- \\
$D$-/$f$-statistics (ABBA-BABA, $f_{2/3/4}$) & \no & \texttt{patterson\_d}, \dots & --- \\
Distance, PCoA, PCA & \no & \texttt{pca}, \texttt{pcoa}, \dots & --- \\
\bottomrule
\end{tabularx}
\vspace{4pt}

\footnotesize{s-a $=$ scikit-allel. \yes\ $=$ output validated (exact or within
machine epsilon) against the named tool; ``VCFtools''/``partial'' indicate the
reference used or partial coverage. $^{\dag}$Watterson's $\theta$ is reported
as a component of the \texttt{tajima-d} windowed output. Group B statistics
require the whole genotype/haplotype matrix in memory and are outside
VariantFlow's streaming scope; users should use scikit-allel for these.}
\end{table}

% ======================================================================
\newpage
\section*{Note S1: Missing-data-aware $\pi$ and $d_{XY}$}

VariantFlow's \texttt{pixy} subcommand consumes an all-sites VCF and, at each
site, tallies pairwise allelic differences and comparisons over non-missing
gametes only: for a population with per-allele counts $c_a$ and
$n=\sum_a c_a$ non-missing gametes, comparisons $=\binom{n}{2}$ and
differences $=(n^2-\sum_a c_a^2)/2$; for a population pair with counts
$c^{(1)}_a,c^{(2)}_a$, comparisons $=n_1 n_2$ and differences
$=n_1 n_2-\sum_a c^{(1)}_a c^{(2)}_a$. Windowed $\pi$ and $d_{XY}$ are the
ratio of summed differences to summed comparisons over all sites in the
window. Because invariant sites contribute comparisons with zero differences
and missing genotypes are excluded per site, the estimator is unbiased under
non-uniform missingness, matching pixy~\citep{pixy2021} exactly (Table~S4)
while running as a compiled selective-execution pass.

% ======================================================================
\newpage
\section*{Note S2: Correctness validation methodology}

Each benchmark row in Table~1 of the main text undergoes correctness
validation before any speedup claim is reported. The validation process:

\begin{enumerate}
\item Run VariantFlow and the baseline tool (bcftools or VCFtools) on
  identical input with equivalent parameters.
\item Normalize outputs: strip headers, sort by genomic position, unify
  column separators.
\item Compute \texttt{diff} on normalized outputs. A row earns
  \checkmark\ only if the diff is empty (zero differing lines).
\item For floating-point statistics (e.g., $F_{\text{MISS}}$, $\pi$),
  additionally verify maximum absolute difference $<10^{-10}$.
\end{enumerate}

For the scikit-allel cross-check, we independently compute per-individual
missingness, allele frequency, and the site frequency spectrum from the
genotype matrix and compare against VariantFlow's output. This provides a
third-party validation independent of both VariantFlow and VCFtools codebases.

\bibliographystyle{apalike}
\bibliography{references}

\end{document}
